---
title: "R Notebook"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---

### Enviornment
1 + 100

2^2

7/2

4-1

2*2

3 + 5*2

(3+5)*2

2e2

2/100000

sin(1)

log(1)

exp(0)

1 == 1

1 >= 1

1 < 1

1 != 2


x <- 10

x <- 10

x

x + x

x - 5

x^2

y <- x + x

y

1:5

2^(1:5)

v <- 1:5

2^v

log10(v)

```{r}


```

## Package managment

```{r}
installed.packages()
```
```{r}
#install.packages("vegan", dependencies = TRUE)
```

```{r}
#library(vegan)
```

```{r}
#remove.packages("vegan")
```

## Getting Help
```{r}
?plot()
```
```{r}
help(plot)
```
```{r}
?"<-"
```
```{r}
vignette("FAQ-vegan")
```

```{r}

```


## Data and their formats

```{r}
cats <- data.frame(coat = c("calico", "black", "taffy"),
                   weight = c(2.1,5,3.2),
                   likes_string = c(1,0,1))
```

```{r}
cats
```

```{r}
getwd()
```
```{r}
write.csv(cats, file = "./cats.csv")
```

```{r}
#cats <- read.csv("./data/cats/csv")
```

```{r}
cats$coat
```
```{r}
cats$weight
```

```{r}
cats$weight*10
```
```{r}
log(cats$weight)
```
```{r}
logweight <- log(cats$weight)
cbind(cats,logweight)
```
```{r}
cats <- cbind(cats,logweight)
```

```{r}
cats
```

```{r}
paste("My cat is ", cats$coat, ", and it weighs ", cats$weight, " kg.", sep="")
```

### Data types

```{r}
typeof(cats$coat)
```

```{r}
typeof(cats$weight)
```

```{r}
typeof(cats$likes_string)
```

```{r}
typeof(3.14159)
```
```{r}
typeof(FALSE)
```
```{r}
typeof(TRUE)
```
```{r}
typeof(T)
typeof(F)
```
```{r}
typeof(cats)
```
```{r}
class(cats)
```
```{r}
file.show("./data/cats.csv")
```

```{r}
#cats$likes_string <- as.logical(cats$likes_string)
#type
```
```{r}
#cats$likes_string
```
```{r}
c("a","b")
```

```{r}
ab <- c("a", "b")
ab
```

```{r}
c(ab,"c")
```

```{r}
c(ab, ab)
```

```{r}
1:10
```

```{r}
seq(10)
```

```{r}
z <- seq(10)
head(z, n=3)
```
```{r}
length(z)
```

```{r}
class(z)
```

```{r}
typeof(z)
```

```{r}
seq(0,100, by=5)
```


### Data Frames

```{r}
cats
```

```{r}
str(cats$coat)
```

```{r}
coats <- c("tabby", "tortoiseshell", "tortoiseshell", "black", "tabby")
coats
```
```{r}
str(coats)
```
```{r}
factor(coats)
```
```{r}
categories <- factor(coats)
class(coats)
class(categories)
```
```{r}
str(categories)
```

### Lists

```{r}
list_example <- list(title="Numbers", numbers = 1:10, data=TRUE)
```
```{r}
list_example
```
```{r}
another_list <- list(1, "a",TRUE, 1+1i)
another_list
```

```{r}
typeof(list_example)
class(list_example)
```
```{r}
typeof(cats)
class(cats)
```

```{r}
data.frame(list_example)
```
```{r}
cats
```
```{r}
cats[,2]
```

```{r}
cats[3,]
```

```{r}
cats[2:3,c(1,3)]
```

### Matricies

```{r}
matrix_example <- matrix(0, ncol=5, nrow=3)
```

```{r}
matrix_example
```


```{r}
class(matrix_example)
```
```{r}
typeof(matrix_example)
```
```{r}
str(matrix_example)
```
```{r}
dim(matrix_example)
```
```{r}
ncol(matrix_example)
```
```{r}
nrow(matrix_example)
```
```{r}
class(data.frame(matrix_example))
```
```{r}
df_example <- data.frame(matrix_example)
df_example
```

### Subsetting

```{r}
p <- c(2.3,6.9,4.0,24.3,2)
p
```
```{r}
names(p) <- c('a','b','c','d','e')
p
```
```{r}
p[2:4]
```
```{r}
p[c(1,5)]
```
```{r}
p[c(1,1,1,3,5,5)]
```
```{r}
p[6]
```
```{r}
p[-3]
```
```{r}
p[c(-1,-5)]
```
```{r}
p[-(2:4)]
```
```{r}
p[c('a','c')]
```
```{r}
p[c(T,F,T,F,T)]
```
```{r}
p[names(p) != 'c']
```

### Factors

```{r}
f <- factor(c('a','b','c','d','e'))
f
```

```{r}
f[f == 'a']
```

```{r}
f[1:3]
```
```{r}
f[f %in% c('b','c')]
```

```{r}
f[-3]
```

```{r}
f2 <- c('a','a','b','c','c')
f2
```
```{r}
f2[f2 == 'a']
```
```{r}
f2[f2 %in% c('a','c')]
```

### Matricies Resumed

```{r}
set.seed(21)
m <- matrix(rnorm(6*4), ncol=4, nrow=6)
m
```

```{r}
m[3:4, c(3,1)]
```
```{r}
m[,c(3,4)]
```

### Lists revisted

```{r}
xlist <- list(a="BIOL48006220",b=seq(1,10,by = 0.5), data="Grade")
xlist
```

```{r}
xlist[1:2]
```

```{r}
xlist['a']
```

```{r}
xlist[['a']]
```

```{r}
xlist['b']
```

```{r}
xlist[['b']]
```

```{r}
xlist$b
```

### Data Frames Revisited 

```{r}
gp <- read.csv('forest_area_sq_km.csv')
```

```{r}
head(gp, n=10L)
```
```{r}
head(gp[3], n=10L)
```

```{r}
nrow(gp)
```

```{r}
head(gp[["country"]], 10L)
```

```{r}
gp$X2001
```

```{r}
gp[c(1,3), 2:5]
```

```{r}
gp[ which(gp$X1989 <= 100 & gp$X2019 >= 100),]
```

## Conditionals and Flow

```{r}
n <- 10

if (n < 10) {
  print("n is less than 10")
} else if (n > 10) {
  print("n is greater than 10")
} else {
  print("n is equal to 10")
}
```

```{r}
for (n in seq(0,20)) {
  if (n < 10) {
    print("n is less than 10")
  } else if (n > 10) {
    print("n is greater than 10")
  } else {
    print("n is equal to 10")
  }
}  
```
```{r}
g <- 0 

while (g <=10) {
  print(paste(g,"is less than or equal to 10"))
  g <- g+1
}
```

## Plotting and Figures

```{r}
#install.packages("ggplot2")
```

```{r}
library(ggplot2)
```

```{r}
forestdata <- read.csv('forest_area_sq_km.csv')

head(forestdata)
```

```{r}
ggplot(data=forestdata, mapping=aes(x=X1989, y=X2019)) + geom_point()
```

```{r}
gp <- read.csv('gapminder_all.csv')
gp
```
```{r}
ggplot(data=gp, mapping=aes(x=gdpPercap_1952, y=pop_1952)) + geom_point()
```

```{r}
ggplot(data=gp, mapping=aes(x=gdpPercap_2002, y=pop_2002)) + geom_point()
```

```{r}
gapminder <- read.csv("https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/gh-pages/_episodes_rmd/data/gapminder_data.csv")
head(gapminder)
```

```{r}
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp)) + geom_point()
```

```{r}
ggplot(data=gapminder, mapping=aes(x=year, y=lifeExp, by=country, color=continent)) + geom_line()
```

```{r}
ggplot(data=gapminder, mapping=aes(x=year, y=lifeExp, by=country)) + geom_line(mapping = aes(color=continent)) + geom_point()
```

```{r}
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp)) + geom_point()
```

```{r}
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp)) + geom_point(alpha=0.25) + scale_x_log10()
```

```{r}
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp)) + geom_point(alpha=0.25, color="purple", size = 0.9, pch = 9) + scale_x_log10() + geom_smooth(method=lm, color="gold", size=1.25)
```

```{r}
africas <- gapminder[gapminder$continent == "Africa", ]
head(africas)
```

```{r}
ggplot(data=africas, mapping = aes(x=year, y=lifeExp)) + 
  geom_line(color = 'blue') + 
  facet_wrap( ~ country) +
  theme(axis.text.x = element_text(angle = 90)) +
  labs( 
    x = "year",
    y = "life Expentency",
    title = "Life Expectency Over Time in African Countries"
  )
```

```{r}
AfricanLifeExp <- ggplot(data=africas, mapping = aes(x=year, y=lifeExp)) + 
  geom_line(color = 'blue') + 
  facet_wrap( ~ country) +
  theme(axis.text.x = element_text(angle = 90)) +
  labs( 
    x = "year",
    y = "life Expentency",
    title = "Life Expectency Over Time in African Countries"
  )
```

```{r}
ggsave(filename = "AfricanLifeExp.png", plot = AfricanLifeExp, width = 24, height = 40, dpi = 300, units = "cm")
```

```{r}
pdf(file = "AfricanLifeExp.pdf", width = 24, height = 40)
plot(AfricanLifeExp)
dev.off()
```

```{r}
write.table(gapminder, file="gapminder_web.csv", sep = ",")
```

```{r}
write.csv(africas, file="gapminder_web_africas.csv")
```

## Statistics 

### Additional ways of importing/reading data

```{r}
rand <- c(12, 54, 33, 98, 65)
rand
```

```{r}
sum(rand)
```

```{r}
length(rand)
```

```{r}
avg.rand <- sum(rand)/length(rand)
avg.rand
```

```{r}
mean(rand)
```

```{r}
sort(rand)
```

```{r}
min(rand)
max(rand)
```

```{r}
cumsum(rand)
```

```{r}
diff(rand)
```

```{r}
rand[2]
```

```{r}
rand*2
```

```{r}
pedes <- scan()
```

```{r}
pedes
```

### Analyses

#### univariate statistics

1. Catoegorical Data
  a. Barplots
```{r}
beer <- c(3, 4, 1, 1, 3, 4, 3, 3, 1, 3, 2, 1, 2, 1, 2, 3, 2, 3, 1, 1, 1, 1, 4, 3, 1)
length(beer)
```

```{r}
barplot(beer)
```



```{r}
barplot(table(beer), xlab="Beer", ylab="frequency") # Counts
```

```{r}
barplot(table(beer)/length(beer), xlab="Beer", ylab="Proportion") # Proportions instead of frequencies
```

  b. Pie Charts
```{r}
pie(table(beer), main="Beer preferrence by students")
```

2. Numerical Data
  a. Stem-and-lead Plots
```{r}
stem(pedes)
```

  b. Strip
```{r}
stripchart(pedes, method="stack")
```

3. Measures of Center
  a. Mean
```{r}
mean(pedes)
```

```{r}
mean(gapminder2$lifeExp)
```

  b. Median 
```{r}
median(pedes)
```

```{r}
median(gapminder$lifeExp)
```

```{r}
with(gapminder, median(lifeExp))
```

  c. Mode
```{r}
which(table(pedes) == max(tbale(pedes)))
```

4. Variation
  a. Range
```{r}
range(pedes)
```

```{r}
diff(range(pedes))
```

  b. variance
```{r}
var(pedes)
```


```{r}
sd(pedes)
```

  C. IQRs
  
```{r}
IQR(pedes)
```

```{r}
scale(pedes)
```

  e. summary
```{r}
summary(pedes)
```


5. Plot
  a. histograms
```{r}
hist(pedes, breaks="scott")
```

```{r}
hist(pedes, breaks+"scott", prob=TRUE)
```

```{r}
hist(pedes, breaks="scotts", prob=T)

```

```{r}
plot(density(pedes))
```

  b. Box plots
```{r}
boxplot(pedes)
```

```{r}
summary(boxplot(pedes))
```

#### Bivariate and multivariant 

```{r}
spid.gen <- read.csv("data/spider_genitalia.csv", header = T)
spid.gen
```

```{r}
bxoplot(spid.gen$left.bulb ~ spid.gen$right.bulb)
```

```{r}
plot(spid.gen$left.bulb ~ spid.gen$right.bulb)
```

```{r}
plot(spid.gen$left.bulb ~ spid.gen$right.bulb, pch = as.character(spid.gen$habitat))
```

  C. Linear regression
```{r}
gen.reg <- lm(spid.gen$left.bulb ~ spid.gen$right.bulb)
gen.reg
```

```{r}
summary(gen.reg)
```

```{r}
plot(spid.gen$left.bulb ~ spid.gen$right.bulb, pch + as.character(spid.gen$habitat))
abline(gen.reg)
```

  d. Correlation Coefficients & Soearman Rank Correlation

```{r}
cor.gen <- with(spid.gen, cor(left.bulb, right.bulb))
cor.gen^2
```

```{r}
spearman.cor.gen <- with(spid.gen, cor(left.bulb, right.bulb, methof"spearman"))
spearman.cor.gen^2
```

  e. residuals 
```{r}
residuals(gen.reg)
```

  f. transformation
```{r}
plot(spid.gen$left.bulb^2~spid.gen$right.bulb)
```

  2. Comparing Discrete Treatment Effects
    a. chi-squared

```{r}
obs_weighted <- c(4,15,6,15,18,2)
obs_fair <- c(10,10,10,10,10,10)
exp <- c(.16,.17,.16,.17,.17,.17)
chisq.test(obs_weighted, p=exp)
chisq.test(oobs_fair, p=exp)
```

  b. T-tests
```{r}
#H0 - true mean is equal to 0, but you can set true mean to other values with mu=
t.test(spid.gen$carapace.length)

t.test(spid.gen$carapace.length, mu=29)
```

```{r}
habitat.t2 <- t.test(spid.gen$carapace.length~spid.gen$habitat)
habitat.t2
```

```{r}
habitat.tless <- t.test(spid.gen$left.bulb, alternative = "less")
habitat.tless
```


```{r}
habitat.tgreater <- t.test(spid.gen$left.bulb, alternative = "greater")
habitat.tgreater
```

```{r}
plot(1:25, rep(0.25,25), pch=1:25, col=1:25, ylim=c(0,6), cex=2, ylab="Line types (lty) 1 to 6", xlab="Plotting character (pch) 1 to 25 and colours (col) 1 to 8", main="Line types (lty), plotting characters (pch), \nand colors (col) for plot and xyplot", lab=c(25,7,2))

points(1:8, rep(0.5,8), pch=20, col=1:8, cex=4)

abline(h=1:6, lty=1:6, col=1:6, lwd=3)
```

```{r}
pdf("./DATA/PlotCheatSheet.pdf")

plot(1:25, rep(0.25,25), pch=1:25, col=1:25, ylim=c(0,6), cex=2, ylab="Line types (lty) 1 to 6", xlab="Plotting character (pch) 1 to 25 and colours (col) 1 to 8", main="Line types (lty), plotting characters (pch), \nand colors (col) for plot and xyplot", lab=c(25,7,2))

points(1:8, rep(0.5,8), pch=20, col=1:8, cex=4)

abline(h=1:6, lty=1:6, col=1:6, lwd=3)

dev.off()
```

## Looping and Conditionals

a. If then statments

```{r}
x <- 0
if(x > 0){
  print("positive number")
} else if (x < 0) {
  print("negativce number")
} else 
  print("zero")
```

```{r}
a <- c(5,7,2,9)
ifelse(a %% 2 == 0, "even", "odd")
```

b. For loops

```{r}
z <- c(2,5,3,7,8,9,11,7)
count <- 0 

for (val in z) {
  if(val %% 2 == 0) count = count+1
}
print(count)
```

c. while statement
```{r}
i <- 1
while (i < 5) {
  print(i)
  i = i+1
}
```

d. Interruptions

```{r}
x <- 1:5

for (val in x) {
  if (val == 3) {
    break
  }
  print(val)
}
```

```{r}
x <- 1:5
for (val in x) {
  if (val == 3) {
    next
  }
  print(val)
}
```

```{r}
x <- 1:5
repeat {
  print(x)
  x = x+1
}
```

## randomization and dataset management (short)

a. dataset managment
```{r}
B <- matrix(
  c(2,3,4,1,5,98), 
  nrow=3,
  ncol=2
)

B
```

```{r}
t(B)
```

```{r}
C <- matrix(
  c(7,4,2),
  nrow = 3,
  ncol = 1
)

C
```

```{r}
BC <- cbind(B,C)

BC
```

```{r}
c(B)
```

b. Ranomizations

```{r}
# rnorm(n, mean, sd)
# sample(x, size, replace = FALSE, prob = NULL)
```

```{r}
ndist <- rnorm(100, 50, 10)
ndist
```

```{r}
mean(ndist)
sd(ndist)
```

```{r}
hist(ndist)
```


```{r}
sample(ndist, 10)
```

```{r}
test <- 1:100
test
```

```{r}
sample(test, 50, replace = FALSE)
```

```{r}
sample(test, 50, replace = TRUE)
```

